---
title: "Homework 2"
subtitle: "DATA 624"
author: "Kevin Havis"
date: today
date-format: "MMMM D, YYYY"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
execute:
  warning: false
  message: false
  echo: true
  cache: false
---

```{r}
#| label: setup
#| include: false

# Load required libraries
library(tidyverse)
library(fpp3)
library(ggplot2)
library(forcats)
library(slider)


# Set ggplot theme
theme_set(theme_minimal())
```

# Homework 2 - Chapter 3 Exercises

## Exercise 3.1
Consider the GDP information in `global_economy`. Plot the GDP per capita for each country over time. Which country has the highest GDP per capita? How has this changed over time?

```{r}
head(tsibbledata::global_economy)

# There's 250+ countries so limiting for plotting purposes
autoplot(global_economy, .vars=gdp_capita)

```

## Exercise 3.2
For each of the following series, make a graph of the data. If transforming seems appropriate, do so and describe the effect.
* United States GDP from `global_economy`.
* Slaughter of Victorian "Bulls, bullocks and steers" in `aus_livestock`.
* Victorian Electricity Demand from `vic_elec`.
* Gas production from `aus_production`.

### Response

#### U.S. Economy

The U.S. Economy data appeared to be slightly quadratic, so I applied a square root on the GDP to get a more linear result. We also could a apply a per-capita effect if desired, useful when comparing to other countries.

```{r}
# US Econ
us_econ <- global_economy |> 
  filter(Code == "USA")

autoplot(us_econ) + labs(title = "United States Economy")
autoplot(us_econ, sqrt(GDP)) + labs(title="United States Economy w/ Square Root GDP")
```


#### Beef Livestock

This timeseries is quite busy with many month-to-month variations. I have applied a simple five month moving average to help smooth the overal pattern, which still reveals seasonality, but makes it more clear and the trend more obvious.

```{r}
# Livestock
beef <- aus_livestock |> 
  filter(Animal == "Bulls, bullocks and steers") |> 
  filter(State == "Victoria")

# A very noisy time series
autoplot(beef) + labs(title="Victorian Bulls, bullocks, and steers (Original)")

# Use basic moving average smoothing
smooth_beef <- beef |> 
  arrange(Month) |> 
  mutate(
    ma_5 = slide_dbl(Count, mean, .before=2, .after=2, .complete=TRUE)
  )
autoplot(smooth_beef, ma_5) + labs(title="Victorian Bulls, bullocks, and steers (Moving Average)")
```
#### Victoria Electricity Demand

This is the most detailed data we have seen yet; the data represents electricity demand down to the half hour.

While there are likely interesting observations in such granular data, we could use a moving average of the demand to eliminate the interday seasonality, looking back and forward by 12 hours (24 half hour intervals).

While still a very visually dense graph, this transformation it "thins" the series enough to be more digestible.

These types of transformations demand on the objective, and instead of transforming, decomposition would be more generally useful to extract different levels of seasonality.


```{r}
# Vic Elec Demand
vic_elec |> 
  autoplot() + labs(title="Victoria Electricity Demand")

vic_elec |> 
  arrange(Time) |> 
  mutate(ma_24 = slide_dbl(Demand, mean, .before =24, .after=24, .complete=TRUE)) |> 
  autoplot(ma_24) + labs(title="Victoria Electricity Demand (Daily Moving Average)")

```

#### AUS Production

Australian gas production has increased but become more varied as time has gone on. This is a great use case for a Box Cox transformation to help "squish" the variance to a regularized amount to make the time series more comparable over time.

We can use the `guerrero` optimizer to find a value of lambda (the Box Cox parameter) that works well for our data.

Upon transforming, we can see our time series has a much more consistent level of variance over time, making further analysis easier. We must remember to reverse the transformation before interpreting though!

```{r}
# AUS Production
aus_production |> 
  autoplot(Gas)

# Straight from the textbook
lambda <- aus_production |>
  features(Gas, features = guerrero) |>
  pull(lambda_guerrero)
aus_production |>
  autoplot(box_cox(Gas, lambda)) +
  labs(title="AUS Gas Production (Guerrero-Optimized Box Cox Transformation")
```


## Exercise 3.3
Why is a Box-Cox transformation unhelpful for the `canadian_gas` data?

### Response

Box Cox transformations can be helpful when the variance of a series trends in one direction or another with some consistency. 

Reviewing the `canadian_gas` series, the variance starts small, increases through the middle of the series, and then becomes small yet more erratic in the recent data. Applying a Box Cox transformation would apply the same effect (compressing or expanding) to *all points* of the time series, meaning the transformation would do a poor job normalizing this series.

A better option would be Loess smoothing, which uses localized data to dynamically smooth the curve.

```{r}
canadian_gas |> 
  autoplot() + labs(title="CAD Gas Production")

gas_lambda <- canadian_gas |>
  features(Volume, features = guerrero) |>
  pull(lambda_guerrero)

canadian_gas |>
  autoplot(box_cox(Volume, gas_lambda)) +
  labs(title="CAD Gas Production (Ineffective Box Cox Transformation")

canadian_gas |> 
  autoplot() +
  geom_smooth() +
  labs(title="CAD Gas Production with Loess Smoothing")
```

## Exercise 3.4
What Box-Cox transformation would you select for your retail data (from Exercise 7 in Section 2.10)?

### Response

The variance of this time series grows quite quickly, and as such we will need a more intense transformation.

A Box Cox transformation with $\lambda = 0$ gives us the desired effect. This is a log transformation, and once applied, we can we see that it has effectively standardized the variance across the series.

```{r}
set.seed(42)

myseries <- aus_retail |>
  filter(`Series ID` == sample(aus_retail$`Series ID`,1))

autoplot(myseries) + labs(title="AUS Retail Turnover")

autoplot(myseries, box_cox(Turnover, 0)) + labs(title="AUS Retail Turnover (Box Cox Log Transformed)")

```

## Exercise 3.5

For the following series, find an appropriate Box-Cox transformation in order to stabilise the variance.
* Tobacco from `aus_production`
* Economy class passengers between Melbourne and Sydney from `ansett`
* Pedestrian counts at Southern Cross Station from `pedestrian`

```{r}

```

## Exercise 3.6
Show that a 3×5 MA is equivalent to a 7-term weighted moving average with weights of 0.067, 0.133, 0.200, 0.200, 0.200, 0.133, and 0.067.
```{r}

```

## Exercise 3.7
Consider the last five years of the Gas data from `aus_production`.
```{r}
gas <- tail(aus_production, 5*4) |> select(Gas)
```

1. Plot the time series. Can you identify seasonal fluctuations and/or a trend-cycle?
2. Use `classical_decomposition` with `type=multiplicative` to calculate the trend-cycle and seasonal indices.
3. Do the results support the graphical interpretation from part a?
4. Compute and plot the seasonally adjusted data.
5. Change one observation to be an outlier (e.g., add 300 to one observation), and recompute the seasonally adjusted data. What is the effect of the outlier?
6. Does it make any difference if the outlier is near the end rather than in the middle of the time series?
```{r}

```

## Exercise 3.8
Recall your retail time series data (from Exercise 7 in Section 2.10). Decompose the series using X-11. Does it reveal any outliers, or unusual features that you had not noticed previously?
```{r}

```

## Exercise 3.9
Figures 3.19 and 3.20 show the result of decomposing the number of persons in the civilian labour force in Australia each month from February 1978 to August 1995.
```{r}

```


