---
title: "Homework 4"
subtitle: "DATA 624"
author: "Kevin Havis"
date: today
date-format: "MMMM D, YYYY"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
execute:
  warning: false
  message: false
  echo: true
  cache: false
---

```{r}
#| label: setup
#| include: false

# Load required libraries
library(tidyverse)
library(mlbench)
library(ggplot2)
library(GGally)
library(corrr)

# Set default chunk options
knitr::opts_chunk$set(
  fig.width = 10,
  fig.height = 8,
  fig.align = "center",
  out.width = "80%"
)

# Set ggplot theme
theme_set(theme_minimal())
```

# Chapter 3 Kuhn, Johnson

## 3.1

```{r}
data <- data(Glass)
```

**Using visualizations, explore the predictor variables to understand their distributions as well as the relationships between predictors.**

We can set up a pair plot to visualize the relationships between the numerical predictors. Below are two versions; a detailed pairplot, and a correlation heatmap.

From these we can quickly highlight a few correlated predictors;

- `Ca` and `RI`: High positive correlation of 81%
- `Si` and `RI`: High negative correlation of 54%
- `Mg`, `Ba`, `Ca`: These three negatively correlate with each other, around 44-50%

```{r}
Glass |> 
  select(where(is.numeric)) |> 
  ggpairs(lower = list(continuous = "points"),
          diag = list(
            continuous = wrap("barDiag",
            bins=ceiling(nrow(Glass)/10))
            ),
          upper = list(continuous = "cor"),
          progress = FALSE)
```

```{r}
# Heatmap

Glass |> 
  select(where(is.numeric)) |> 
  cor() |> 
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") %>%
  mutate(var1 = as.character(var1),
         var2 = as.character(var2)) %>%
  ggplot(aes(x = var1, y = var2, fill = correlation)) +
  geom_tile() +
  labs(x = NULL, y = NULL) +
  scale_fill_distiller(type = "div", palette = "RdBu", 
                       limits = c(-1, 1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Do there appear to be any outliers in the data? Are any predictors skewed?

- `K` has a few very high outliers
- `Ca`, `Ba`, and`Fe` have very long right-tails and skew
  - `Ba` and `Fe` especially have many zero values - notably, they are zero floats and not null values
- `Mg` has several very low outliers (zero or nearly zero)


Are there any relevant transformations of one of more predictors that might improve the classification model?

```{r}
Glass |> 
  ggplot(aes(x=Ca)) + 
  geom_histogram(bins = 100)
  
```


## 3.2

```{r}
data("Soybean")
```

The soybean data can also be found at the UC Irvine machine learning repository. Data were collected to predict disease in 683 soybeans. The 345 predictors are mostly categorical and include information on the environmental conditions (e.g., temperature, precipitation) and plant conditions (e.g., left spots, mold growth). The outcome labels consists of 19 distinct classes.  

Investigate the frequency distributions for the categorical predictors. Are any of the distributions degenerate in the ways discussed earlier in this chapter?

```{r}

```


Roughly 18%% of the data re missing are there any particular predictors that are more likely to be missing? Is the pattern of missing data related to the classes?

```{r}

```


Develop a strategy for handling missing data, either by eliminating predictors or imputation.

```{r}

```


# Appendix

```{r}
#| label: session-info
#| echo: false
sessionInfo()
```