---
title: "Project 1"
subtitle: "DATA 624"
author: "Kevin Havis"
date: today
date-format: "MMMM D, YYYY"
format: 
  pdf:
    toc: false
    number-sections: false
    colorlinks: true
  html:
    toc: false
    toc-depth: 3
    number-sections: false
    code-fold: false
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
  cache: false
---

```{r}
#| label: setup
#| include: false

# Load required libraries
library(tidyverse)
library(fpp3)
library(knitr)
#library(urca)

# Set default chunk options
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  out.width = "80%"
)

# Set ggplot theme
theme_set(theme_minimal())
```

# Project 1

## Part A - ATM Forecast

### Exploratory Data Analysis

Examining the data, I see several entries with missing values at the end of the time series. Assuming these are the intended forecast periods, I will remove them from the raw data.

Plotting the remaining series, we have four ATMs. Of particular note is "ATM 4", which seems to generally have much higher levels than the others, including a massive outlier.


```{r}
# Read in ATM data
atm_data <- readxl::read_excel("ATM624Data.xlsx") |> 
  mutate(DATE = as.Date(DATE, origin = "1899-12-30")) |>
  as_tsibble(index=DATE, key=ATM) |> 
  drop_na()

autoplot(atm_data)

```

We temporarily remove ATM 4 in order to more clearly see the series of the remaining three.

ATM 3 also appears to be unusual, with zero balances through most the time series, then a sudden spike near the end.

ATM 1 and 2 demonstrate large variances of cash balances through the series, presumably showing the ATM's cash reserves gradually being drawn down, then suddenly replenished.

```{r}
autoplot(atm_data |> filter(ATM != "ATM4"))
```

Given the differences, and assuming there's real business reasons behind them, we will analyze and model each ATM separately.

### ATM 1

- Variance is stable
- High seasonality
- No real trend to speak of
- Gaps in time series, need to fill
  - Since the data is very seasonal, use snaive to fill gaps
   - Use 7 days based on ACF plot


```{r}
#| label: Preprocessing and imputation

# Get ATM 1 data
atm1 <- atm_data |>
  filter(ATM == "ATM1") 

# Impute simple mean so we can check ACF
atm1_fillmean <- atm1 |> 
  fill_gaps(Cash = mean(Cash, na.rm = TRUE)) 
  
# Check what seasonality to impute on
atm1_fillmean |> 
  ACF(Cash) |> 
  autoplot() +
  scale_x_continuous(breaks = seq(0, 50, by = 1)) +
  labs(
    title = "ATM 1 ACF w/ Simple Mean Imputation",
    y = "Cash Balance",
    x = "Date"
  )

# Fill gaps using seasonal naive lag = 7
# We do this manually because snaive can't fit with gaps
atm1_filled <- atm1 %>%
  fill_gaps() %>%
  group_by_key() %>%
  mutate(Cash = if_else(is.na(Cash), lag(Cash, 7), Cash)) %>%
  ungroup()

# Confirm gaps are fixed
scan_gaps(atm1_filled) # should be empty
```

 - We select several models that can deal with the seasonality
 - ARIMA has the best MASE for the point estimate
 

```{r}
atm1_fit <- atm1_filled |> 
  #mutate(Cash = difference(Cash, lag=7)) |>
  model(arima = ARIMA(Cash ~ PDQ(period = 7)),
        snaive = SNAIVE(Cash ~ lag("week")),
        snaive_drift = SNAIVE(Cash ~ drift()),
        ets = ETS(Cash))

atm1_fit |> 
  select(arima) |>
  gg_tsresiduals() + labs(title="ARIMA")

atm1_fit |>
  select(snaive) |> 
  gg_tsresiduals() + labs(title="SNAIVE")

atm1_fit |>
  select(snaive_drift) |> 
  gg_tsresiduals() + labs(title="Drift")

atm1_fit |>
  select(ets) |> 
  gg_tsresiduals() + labs(title="ETS")

atm1_fit |> 
  accuracy() |> 
  kable(caption="ATM 1 Point Estimate Accur acy")
```
 - Forecasting out 14 days, training on ~90% of the data
 - ARIMA has the best point forecast, but ETS has the best MASE on the test data
 - We should prefer ETS due to simplicity and explainability of the model, which makes sense given how consistent the series is
 

```{r}
# Split the data into train and test
atm1_train <- atm1_filled |> filter_index(. ~ "2010-03-31")
atm1_test <- atm1_filled |> filter_index("2010-04-01" ~ .)

# Refit models on training data
atm1_fit_train <- atm1_train |>
  model(arima = ARIMA(Cash),
        snaive = SNAIVE(Cash ~ lag("week")),
        snaive_drift = SNAIVE(Cash ~ drift()),
        ets = ETS(Cash))

# Forecast on test data
atm1_fc <- atm1_fit_train |>
  forecast(h = "14 days")

# Plot forecasts against test data
atm1_fc |>
  autoplot(atm1_test) +
  labs(title = "ATM 1 Forecast vs Actuals",
       y = "Cash Balance",
       x = "Date")

# Check accuracy on test data
atm1_fc |>
  accuracy(bind_rows(atm1_train, atm1_test)) |> 
  kable(caption="ATM 1 Test Accuracy")
```

### ATM 2

- We'll perform similar steps for ATM 2
- Using `imputeTS` library going forward for missing dates
- Can't use MPE or MAPE due to zero levels for some observations
- ARIMA again has best point accuracy
- ETS and ARIMA perform well, but ETS performs slightly better and is simpler


```{r}
library(imputeTS)

atm_data <- atm_data |> 
  group_by(ATM) |> 
  fill_gaps() |> 
  mutate(Cash = na_interpolation(Cash, option = "linear")) |> 
  ungroup()
```


```{r}
#| label: Preprocessing and imputation ATM2
# Get ATM 2 data
atm2 <- atm_data |>
  filter(ATM == "ATM2")

autoplot(atm2) +
  labs(title="ATM 2 After Linear Interpolation")

atm2_fit <- atm2 |> 
  model(
    arima = ARIMA(Cash),
    snaive = SNAIVE(Cash ~ lag("week")),
    snaive_drift = SNAIVE(Cash ~ drift()),
    ets = ETS(Cash)
  )

atm2_fit |> 
  select(arima) |>
  gg_tsresiduals() + labs(title = "ARIMA - ATM2")

atm2_fit |>
  select(snaive) |> 
  gg_tsresiduals() + labs(title = "SNAIVE - ATM2")

atm2_fit |>
  select(snaive_drift) |> 
  gg_tsresiduals() + labs(title = "Drift - ATM2")

atm2_fit |>
  select(ets) |> 
  gg_tsresiduals() + labs(title = "ETS - ATM2")

atm2_fit |> 
  accuracy() |> 
  kable(caption="ATM 2 Point Estimate Accuracy")
```
```{r}
# Split the data into train and test
atm2_train <- atm2_filled |> filter_index(. ~ "2010-03-31")
atm2_test <- atm2_filled |> filter_index("2010-04-01" ~ .)

# Refit models on training data
atm2_fit_train <- atm2_train |>
  model(arima = ARIMA(Cash),
        snaive = SNAIVE(Cash ~ lag("week")),
        snaive_drift = SNAIVE(Cash ~ drift()),
        ets = ETS(Cash))

# Forecast on test data
atm2_fc <- atm2_fit_train |>
  forecast(h = "21 days")

# Plot forecasts against test data
atm2_fc |>
  autoplot(atm2_test) +
  labs(title = "ATM 2 Forecast vs Actuals",
       y = "Cash Balance",
       x = "Date")

# Check accuracy on test data
atm2_fc |>
  accuracy(bind_rows(atm2_train, atm2_test)) |> 
  kable(caption="ATM 2 Test Accuracy")
```

### ATM 3

ATM 3 is unique in that it appears to be a new machine, or perhaps recently reinstated, given the data has zero balances until the last three days of the series.

With such limited data, we really can't develop a proper model. From a pure business perspective, we could use the domain knowledge we have about the machine to find a similar machine with more data, and use that to guide our forecasts.

For this exercise, we will simply use a naive random walk to model future balances.


```{r}
#| label: Preprocessing and imputation ATM3
# Get ATM 3 data
atm3 <- atm_data |>
  filter(ATM == "ATM3")

autoplot(atm3)

atm3_fit <- atm3 |> 
  model(RW(Cash))

# Forecast on test data
atm3_fc <- atm3_fit |>
  forecast(h = "7 days")

atm3_fc |>  
  autoplot(atm3) +
  labs(title = "ATM 3 Naive Forecast vs Actuals")
```

### ATM 4

- Significant outlier in a relatively stable series
- We will impute the outlier using linear interpolation

```{r}
# Get ATM 4 data
atm4 <- atm_data |>
  filter(ATM == "ATM4")
autoplot(atm4)

# Find the outlier
atm4 |> 
  slice_max(Cash, n=1)

# Check around the observation
atm4 |> 
  filter_index("2010-02-07" ~ "2010-02-12") |> 
  kable(caption="ATM 4 Outlier on 2010-02-09")

# Impute the outlier using linear interpolation
atm4 <- atm4 |>
  mutate(Cash = if_else(DATE == as.Date("2010-02-09"), NA_real_, Cash)) |> 
  mutate(Cash = na_interpolation(Cash, option = "linear"))

# Check
autoplot(atm4) +
  labs(title="ATM 4 After Outlier Imputation")
```

- With the outlier interpolated, we can proceed to modeling
- For this time series, seasonal naive with drift performs the best

```{r}
# Split data
atm4_train <- atm4 |> filter_index(. ~ "2010-03-31")
atm4_test <- atm4 |> filter_index("2010-04-01" ~ .)

# Fit models on training data
atm4_fit_train <- atm4_train |>
  model(
    arima = ARIMA(Cash),
    snaive = SNAIVE(Cash ~ lag("week")),
    snaive_drift = SNAIVE(Cash ~ drift()),
    ets = ETS(Cash)
  )

# Forecast on test data
atm4_fc <- atm4_fit_train |>
  forecast(h = "21 days")

# Plot forecasts against test data
atm4_fc |>
  autoplot(atm4_test) +
  labs(
    title = "ATM 4 Forecast vs Actuals",
    y = "Cash Balance",
    x = "Date"
  )

# Check accuracy on test data
atm4_fc |>
  accuracy(bind_rows(atm4_train, atm4_test)) |>
  kable(caption = "ATM 4 Test Accuracy")
```


## Part B - Forecasting Power

- The series is relatively consistent except for two outliers
- One outlier is an order of magnitude lower than the rest of the data
  - We can simply raise it by a factor of 10 to impute it
- The other is simply missing
  - Visually it is similar to previous season
  - We should use seasonaly imputation to fill it
- We impute for model training, but should note it in the analysis


```{r}
# Read in power data
power <- readxl::read_excel("ResidentialCustomerForecastLoad-624.xlsx") |> 
  mutate(date = yearmonth(`YYYY-MMM`)) |>
  select(-`YYYY-MMM`) |>
  as_tsibble(index=date)

autoplot(power, KWH)

# Check for gaps in series
scan_gaps(power) # should be empty

# Check for null values
power |> 
  filter(is.na(KWH)) # 2008 Sep is missing

# Check the min outlier
power |> 
  slice_min(KWH, n=1)

# Check around the outlier on July 2010
power |> 
  filter_index("2010-06" ~ "2010-08") |> 
  kable(caption="Power Consumption Around July 2010 Outlier")
```


```{r}
# Impute low outlier by raising it by a factor of 10
power_imp <- power |>
  mutate(KWH = if_else(date == yearmonth("2010 Jul"), KWH * 10, KWH))
autoplot(power_imp, KWH) +
  labs(title="Power Consumption After Outlier Imputation")
```
```{r}
# Impute missing outlier using seasonal naive lag = 12
power_imp <- power_imp |>
  fill_gaps() |>
  mutate(KWH = if_else(is.na(KWH), lag(KWH, 12), KWH))

# Check September 2008 and surrounding obs
power_imp |> 
      filter(date %in% c(
        yearmonth("2007 Sep"), 
        yearmonth("2006 Sep"), 
        yearmonth("2009 Sep")
      )) |>
  kable(caption = "Power Consumption After Missing Value Imputation")

```

- With the outliers handled, we can proceed to modeling
- Very seasonal data so we will try a few seasonal models
- Slight trend upwards overtime

```{r}
power_fit <- power_imp |> 
  model(
    arima = ARIMA(KWH),
    snaive = SNAIVE(KWH ~ lag("year")),
    snaive_drift = SNAIVE(KWH ~ drift()),
    ets = ETS(KWH)
  )

# Forecast through 2014

power_fc <- power_fit |>
  forecast(h = 12)


# Visually Check
power_fc |>  
  autoplot(power_imp) +
  coord_cartesian(xlim = c(yearmonth("2012 Jan"), yearmonth("2014 Dec"))) +
  labs(title = "Power Consumption Forecast vs Actuals",
       y = "KWH",
       x = "Date")
```

- Visually all of these models seem to fit well
- Let's check the accuracy metrics to choose the best one

```{r}
power_fit |> 
  accuracy() |> 
  kable(caption="Power Consumption Point Estimate Accuracy")

power_fit |> 
  select(arima)
```

- ARIMA with (0,0,4) and seasonal component (2,1,0) with drift has been chosen
- This seems reasonable; stationary series, very little autoregressive behavior, but some moving average components to smooth out noise
  - The drift address the slightly upwards trend over time
  - The seasonal component captures the yearly seasonality
- Let's test these models on a train/test split to see if they hold up

```{r}
# Split the data into train and test
power_train <- power_imp |> filter_index(. ~ "2011 Dec")
power_test <- power_imp |> filter_index("2012 Jan" ~ .)
# Refit models on training data
power_fit_train <- power_train |> 
  model(
    arima = ARIMA(KWH),
    snaive = SNAIVE(KWH ~ lag("year")),
    snaive_drift = SNAIVE(KWH ~ drift()),
    ets = ETS(KWH),
  )

# Forecast on test data
power_fc <- power_fit_train |>
  forecast(h = 24)

# Plot forecasts against test data
power_fc |>  autoplot(power_test) +
  labs(title = "Power Consumption Forecast vs Actuals",
       y = "KWH",
       x = "Date")

# Check accuracy
power_fc |>
  accuracy(bind_rows(power_train, power_test)) |> 
  kable(caption="Power Consumption Test Accuracy")
```

- ARIMA still holds out as the best model on the test data
- We would only wish to use another model for simplicity
- Again, we should note the imputation steps taken in the analysis


## Part C - Waterflow Pipes

```{r}
pipe1 <- read_excel("Waterflow_Pipe1.xlsx")
pipe2 <- read_excel("Waterflow_Pipe2.xlsx")
```