---
title: "Project 1"
subtitle: "DATA 624"
author: "Kevin Havis"
date: today
date-format: "MMMM D, YYYY"
format: 
  pdf:
    toc: false
    number-sections: false
    colorlinks: true
  html:
    toc: false
    toc-depth: 3
    number-sections: false
    code-fold: false
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
  cache: false
---

```{r}
#| label: setup
#| include: false

# Load required libraries
library(tidyverse)
library(fpp3)
library(knitr)
#library(urca)

# Set default chunk options
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  out.width = "80%"
)

# Set ggplot theme
theme_set(theme_minimal())
```

# Project 1

## Part A - ATM Forecast

### Exploratory Data Analysis

Examining the data, I see several entries with missing values at the end of the time series. Assuming these are the intended forecast periods, I will remove them from the raw data.

Plotting the remaining series, we have four ATMs. Of particular note is "ATM 4", which seems to generally have much higher levels than the others, including a massive outlier.


```{r}
# Read in ATM data
atm_data <- readxl::read_excel("ATM624Data.xlsx") |> 
  mutate(DATE = as.Date(DATE, origin = "1899-12-30")) |>
  as_tsibble(index=DATE, key=ATM) |> 
  drop_na()

autoplot(atm_data)

```

We temporarily remove ATM 4 in order to more clearly see the series of the remaining three.

ATM 3 also appears to be unusual, with zero balances through most the time series, then a sudden spike near the end.

ATM 1 and 2 demonstrate large variances of cash balances through the series, presumably showing the ATM's cash reserves gradually being drawn down, then suddenly replenished.

```{r}
autoplot(atm_data |> filter(ATM != "ATM4"))
```

Given the differences, and assuming there's real business reasons behind them, we will analyze and model each ATM separately.

### ATM 1

- Variance is stable
- High seasonality
- No real trend to speak of
- Gaps in time series, need to fill
  - Since the data is very seasonal, use snaive to fill gaps
   - Use 7 days based on ACF plot


```{r}
#| label: Preprocessing and imputation

# Get ATM 1 data
atm1 <- atm_data |>
  filter(ATM == "ATM1") 

# Impute simple mean so we can check ACF
atm1_fillmean <- atm1 |> 
  fill_gaps(Cash = mean(Cash, na.rm = TRUE)) 
  
# Check what seasonality to impute on
atm1_fillmean |> 
  ACF(Cash) |> 
  autoplot() +
  scale_x_continuous(breaks = seq(0, 50, by = 1)) +
  labs(
    title = "ATM 1 ACF w/ Simple Mean Imputation",
    y = "Cash Balance",
    x = "Date"
  )

# Fill gaps using seasonal naive lag = 7
# We do this manually because snaive can't fit with gaps
atm1_filled <- atm1 %>%
  fill_gaps() %>%
  group_by_key() %>%
  mutate(Cash = if_else(is.na(Cash), lag(Cash, 7), Cash)) %>%
  ungroup()

# Confirm gaps are fixed
scan_gaps(atm1_filled) # should be empty
```

 - We select several models that can deal with the seasonality
 - ARIMA has the best MASE for the point estimate
 

```{r}
atm1_fit <- atm1_filled |> 
  #mutate(Cash = difference(Cash, lag=7)) |>
  model(arima = ARIMA(Cash ~ PDQ(period = 7)),
        snaive = SNAIVE(Cash ~ lag("week")),
        snaive_drift = SNAIVE(Cash ~ drift()),
        ets = ETS(Cash))

atm1_fit |> 
  select(arima) |>
  gg_tsresiduals() + labs(title="ARIMA")

atm1_fit |>
  select(snaive) |> 
  gg_tsresiduals() + labs(title="SNAIVE")

atm1_fit |>
  select(snaive_drift) |> 
  gg_tsresiduals() + labs(title="Drift")

atm1_fit |>
  select(ets) |> 
  gg_tsresiduals() + labs(title="ETS")

atm1_fit |> 
  accuracy() |> 
  kable(caption="ATM 1 Point Estimate Accur acy")
```
 - Forecasting out 14 days, training on ~90% of the data
 - ARIMA has the best point forecast, but ETS has the best MASE on the test data
 - We should prefer ETS due to simplicity and explainability of the model, which makes sense given how consistent the series is
 

```{r}
# Split the data into train and test
atm1_train <- atm1_filled |> filter_index(. ~ "2010-03-31")
atm1_test <- atm1_filled |> filter_index("2010-04-01" ~ .)

# Refit models on training data
atm1_fit_train <- atm1_train |>
  model(arima = ARIMA(Cash),
        snaive = SNAIVE(Cash ~ lag("week")),
        snaive_drift = SNAIVE(Cash ~ drift()),
        ets = ETS(Cash))

# Forecast on test data
atm1_fc <- atm1_fit_train |>
  forecast(h = "14 days")

# Plot forecasts against test data
atm1_fc |>
  autoplot(atm1_test) +
  labs(title = "ATM 1 Forecast vs Actuals",
       y = "Cash Balance",
       x = "Date")

# Check accuracy on test data
atm1_fc |>
  accuracy(bind_rows(atm1_train, atm1_test)) |> 
  kable(caption="ATM 1 Test Accuracy")
```

### ATM 2

- We'll perform similar steps for ATM 2
- Using `imputeTS` library going forward for missing dates
- Can't use MPE or MAPE due to zero levels for some observations
- ARIMA again has best point accuracy
- ETS and ARIMA perform well, but ETS performs slightly better and is simpler


```{r}
library(imputeTS)

atm_data <- atm_data |> 
  group_by(ATM) |> 
  fill_gaps() |> 
  mutate(Cash = na_interpolation(Cash, option = "linear")) |> 
  ungroup()
```


```{r}
#| label: Preprocessing and imputation ATM2
# Get ATM 2 data
atm2 <- atm_data |>
  filter(ATM == "ATM2")

autoplot(atm2) +
  labs(title="ATM 2 After Linear Interpolation")

atm2_fit <- atm2 |> 
  model(
    arima = ARIMA(Cash),
    snaive = SNAIVE(Cash ~ lag("week")),
    snaive_drift = SNAIVE(Cash ~ drift()),
    ets = ETS(Cash)
  )

atm2_fit |> 
  select(arima) |>
  gg_tsresiduals() + labs(title = "ARIMA - ATM2")

atm2_fit |>
  select(snaive) |> 
  gg_tsresiduals() + labs(title = "SNAIVE - ATM2")

atm2_fit |>
  select(snaive_drift) |> 
  gg_tsresiduals() + labs(title = "Drift - ATM2")

atm2_fit |>
  select(ets) |> 
  gg_tsresiduals() + labs(title = "ETS - ATM2")

atm2_fit |> 
  accuracy() |> 
  kable(caption="ATM 2 Point Estimate Accuracy")
```
```{r}
# Split the data into train and test
atm2_train <- atm2_filled |> filter_index(. ~ "2010-03-31")
atm2_test <- atm2_filled |> filter_index("2010-04-01" ~ .)

# Refit models on training data
atm2_fit_train <- atm2_train |>
  model(arima = ARIMA(Cash),
        snaive = SNAIVE(Cash ~ lag("week")),
        snaive_drift = SNAIVE(Cash ~ drift()),
        ets = ETS(Cash))

# Forecast on test data
atm2_fc <- atm2_fit_train |>
  forecast(h = "21 days")

# Plot forecasts against test data
atm2_fc |>
  autoplot(atm2_test) +
  labs(title = "ATM 2 Forecast vs Actuals",
       y = "Cash Balance",
       x = "Date")

# Check accuracy on test data
atm2_fc |>
  accuracy(bind_rows(atm2_train, atm2_test)) |> 
  kable(caption="ATM 2 Test Accuracy")
```

### ATM 3

ATM 3 is unique in that it appears to be a new machine, or perhaps recently reinstated, given the data has zero balances until the last three days of the series.

With such limited data, we really can't develop a proper model. From a pure business perspective, we could use the domain knowledge we have about the machine to find a similar machine with more data, and use that to guide our forecasts.

For this exercise, we will simply use a naive random walk to model future balances.


```{r}
#| label: Preprocessing and imputation ATM3
# Get ATM 3 data
atm3 <- atm_data |>
  filter(ATM == "ATM3")

autoplot(atm3)

atm3_fit <- atm3 |> 
  model(RW(Cash))

# Forecast on test data
atm3_fc <- atm3_fit |>
  forecast(h = "7 days")

atm3_fc |>  
  autoplot(atm3) +
  labs(title = "ATM 3 Naive Forecast vs Actuals")
```

### ATM 4

- Significant outlier in a relatively stable series
- We will impute the outlier using linear interpolation

```{r}
# Get ATM 4 data
atm4 <- atm_data |>
  filter(ATM == "ATM4")
autoplot(atm4)

# Find the outlier
atm4 |> 
  slice_max(Cash, n=1)

# Check around the observation
atm4 |> 
  filter_index("2010-02-07" ~ "2010-02-12") |> 
  kable(caption="ATM 4 Outlier on 2010-02-09")

# Impute the outlier using linear interpolation
atm4 <- atm4 |>
  mutate(Cash = if_else(DATE == as.Date("2010-02-09"), NA_real_, Cash)) |> 
  mutate(Cash = na_interpolation(Cash, option = "linear"))

# Check
autoplot(atm4) +
  labs(title="ATM 4 After Outlier Imputation")
```

- With the outlier interpolated, we can proceed to modeling
- For this time series, seasonal naive with drift performs the best

```{r}
# Split data
atm4_train <- atm4 |> filter_index(. ~ "2010-03-31")
atm4_test <- atm4 |> filter_index("2010-04-01" ~ .)

# Fit models on training data
atm4_fit_train <- atm4_train |>
  model(
    arima = ARIMA(Cash),
    snaive = SNAIVE(Cash ~ lag("week")),
    snaive_drift = SNAIVE(Cash ~ drift()),
    ets = ETS(Cash)
  )

# Forecast on test data
atm4_fc <- atm4_fit_train |>
  forecast(h = "21 days")

# Plot forecasts against test data
atm4_fc |>
  autoplot(atm4_test) +
  labs(
    title = "ATM 4 Forecast vs Actuals",
    y = "Cash Balance",
    x = "Date"
  )

# Check accuracy on test data
atm4_fc |>
  accuracy(bind_rows(atm4_train, atm4_test)) |>
  kable(caption = "ATM 4 Test Accuracy")
```


## Part B - Forecasting Power

```{r}
# Read in power data
atm_data <- read_excel("ResidentialCustomerForecastLoad-624.xlsx")


```

## Part C - Waterflow Pipes

```{r}
pipe1 <- read_excel("Waterflow_Pipe1.xlsx")
pipe2 <- read_excel("Waterflow_Pipe2.xlsx")
```