---
title: "Project 1"
subtitle: "DATA 624"
author: "Kevin Havis"
date: today
date-format: "MMMM D, YYYY"
format: 
  pdf:
    toc: false
    number-sections: false
    colorlinks: true
  html:
    toc: false
    toc-depth: 3
    number-sections: false
    code-fold: false
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
  cache: false
---

```{r}
#| label: setup
#| include: false

# Load required libraries
library(tidyverse)
library(fpp3)
library(knitr)

# Set default chunk options
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  out.width = "80%"
)

# Set ggplot theme
theme_set(theme_minimal())
```

# Project 1

## Part A - ATM Forecast

### Exploratory Data Analysis

Examining the data, I see several entries with missing values. Assuming these are the intended forecast periods, I will remove them from the raw data.

Plotting the remaining series, we have four ATMs. Of particular note is "ATM 4", which seems to generally have much higher levels than the others, including a massive outlier.


```{r}
# Read in ATM data
atm_data <- readxl::read_excel("ATM624Data.xlsx") |> 
  as_tsibble(index=DATE, key=ATM) |> 
  drop_na()

autoplot(atm_data)

```

We temporarily remove ATM 4 in order to more clearly see the series of the remaining three.

ATM 3 also appears to be unusual, with zero balances through most the time series, then a sudden spike near the end.

ATM 1 and 2 demonstrate large variances of cash balances through the series, presumably showing the ATM's cash reserves gradually being drawn down, then suddenly replenished.

```{r}
autoplot(atm_data |> filter(ATM != "ATM4"))
```

Given the differences, and assuming there's real business reasons behind them, we will model each ATM separately.

### ATM 1

- Variance is stable
- High seasonality
- No real trend to speak of

```{r}
# Check for nulls in the time series
atm_data |> 
  filter(is.na(DATE))
```

```{r}
atm1 <- atm_data |> filter(ATM == "ATM1")
autoplot(atm1)

atm1_fit <- atm1 |> 
  model(
    mean = MEAN(Cash),
    snaive = SNAIVE(Cash),
    ets = ETS(Cash),
    arima = ARIMA(Cash)
  )
```



## Part B - Forecasting Power

```{r}
# Read in power data
atm_data <- read_excel("ResidentialCustomerForecastLoad-624.xlsx")


```

## Part C - Waterflow Pipes

```{r}
pipe1 <- read_excel("Waterflow_Pipe1.xlsx")
pipe2 <- read_excel("Waterflow_Pipe2.xlsx")
```

