---
title: "ABC Beverage pH Prediction: Technical Report"
author: "Data Science Team"
date: today
format: html
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
library(readxl)
library(tidyverse)
library(randomForest)
library(xgboost)
library(kableExtra)
library(caret)
library(caretEnsemble)
```

## Executive Summary

This report documents the development and validation of a predictive model for pH levels in ABC Beverage's manufacturing process. Multiple modeling approaches were evaluated, and a Random Forest model was selected based on predictive accuracy and interpretability, achieving a $R^2$ of ~69% on cross fold validation training sets.

------------------------------------------------------------------------

## Load Data

```{r}
# Update these paths if necessary
train_data <- read_excel("data/train_data.xlsx")
test_data <- read_excel("data/test_data.xlsx")
test_data <- test_data %>% select(-PH) # Remove empty PH column
```


## EDA

```{r}
eda_summary <- function(data) {
  cat("=== Dataset Overview ===\n")
  cat("Dimensions:", nrow(data), "rows x", ncol(data), "columns\n\n")
  
  cat("=== Missing Values ===\n")
  missing <- colSums(is.na(data))
  if (sum(missing) == 0) {
    cat("No missing values\n\n")
  } else {
    print(missing[missing > 0])
    cat("\n")
  }
  
  cat("=== Numeric Summary ===\n")
  data %>%
    select(where(is.numeric)) %>%
    summary() %>%
    print()
  
  cat("\n=== Target Variable (PH) ===\n")
  cat("Mean:", round(mean(data$PH), 3), "\n")
  cat("SD:", round(sd(data$PH), 3), "\n")
  cat("Range:", round(min(data$PH), 3), "-", round(max(data$PH), 3), "\n")
  
  # Return correlation with target
  cat("\n=== Top Correlations with PH ===\n")
  data %>%
    select(where(is.numeric)) %>%
    cor(use = "complete.obs") %>%
    as_tibble(rownames = "feature") %>%
    select(feature, PH) %>%
    filter(feature != "PH") %>%
    mutate(abs_cor = abs(PH)) %>%
    arrange(desc(abs_cor)) |> 
    kable()
}

eda_summary(train_data)
```


## Data Preprocessing

We found in our experiments that removing missing values was more effective than KNN imputation for all selected models.

```{r} 
clean_data <- function(df) {
  df |> 
    rename_with(~str_replace_all(., " ", "_")) %>%
    drop_na() %>%
    mutate(Brand_Code = as.numeric(as.factor(Brand_Code)))
}

train_data <- clean_data(train_data)
test_data <- clean_data(test_data)
```


------------------------------------------------------------------------

## Modeling

```{r}
# Cross-validation setup
ctrl <- trainControl(
  method = "cv", 
  number = 5, 
  index = createFolds(train_data$PH, k = 5, returnTrain = TRUE)
)

# Train all models with CV
models <- caretList(
  PH ~ ., 
  data = train_data,
  trControl = ctrl,
  tuneList = list(
    lm = caretModelSpec(method = "lm"),
    dt = caretModelSpec(method = "rpart"),
    rf = caretModelSpec(method = "rf", importance = TRUE),
    xgb = caretModelSpec(method = "xgbTree", verbosity = 0)
  )
)

# Get training set metrics
train_metrics <- resamples(models) %>%
    summary() %>%
    .$statistics %>%
    map_df(~ as_tibble(.x, rownames = "model"), .id = "metric") %>%
    filter(metric %in% c("RMSE", "Rsquared", "MAE")) %>%
    select(metric, model, mean = Mean) %>%
    pivot_wider(names_from = metric, values_from = mean)
```


We evaluated three distinct modeling approaches:

### 1. Linear Regression

-   **Method:** Ordinary least squares
-   **RMSE:** \[Value\]
-   **R-Squared:** \[Value\]
-   **Assessment:** Assumes linear relationships; baseline approach

### 2. Decision Tree

-   **Method:** Single tree with pruning
-   **RMSE:** \[Value\]
-   **R-Squared:** \[Value\]
-   **Assessment:** Captures non-linearities but prone to overfitting

### 3. Random Forest

-   **Method:** Ensemble of 500 trees
-   **RMSE:** \[Value\]
-   **R-Squared:** \[Value\]
-   **Assessment:** Best balance of accuracy and generalization

### 4. XGBoost

- **Method**: Gradient boosting with 100 rounds, max depth 6, learning rate 0.1
- **RMSE**: [Value]
- **R-Squared**: [Value]
- **Assessment**: Powerful ensemble method often achieving top performance on tabular data

------------------------------------------------------------------------

## Model Results (Training)

```{r}
train_metrics |> kable()
```

```{r}
models$rf$finalModel$ntree
```


## Model Predictions

```{r}
predict_all_models <- function(models, new_data) {
  tibble(
    lm = predict(models$lm, new_data),
    dt = predict(models$dt, new_data),
    rf = predict(models$rf, new_data),
    xgb = predict(models$xgb, new_data)
  )
}

predictions <- as_tibble(predict(models, test_data))
```

------------------------------------------------------------------------


## Final Model

After evaluating all models, the Random Forest model was selected as the
final model due to its superior performance on training set data, as well as the explainability.


------------------------------------------------------------------------

## Feature Importance Analysis

```{r}
varImp(models$rf)$importance %>%
  as_tibble(rownames = "feature") %>%
  rename(importance = Overall) %>%
  mutate(importance_scaled = 100 * importance / max(importance)) %>%
  arrange(desc(importance_scaled)) |>
  kable()
```

**Key Findings:** - Top 3 features account for approximately \[%\] of
predictive power - Model captures both linear and non-linear
relationships in manufacturing process - \[Feature 1\]: Most important
driver - \[Feature 2\]: Secondary driver - \[Feature 3\]: Tertiary
driver

------------------------------------------------------------------------


## Model Validation

**Cross-Validation Approach:** - Method: 5-fold cross-validation -
Splits: Data divided into 5 equal parts - Process: Model trained on 4
folds, tested on remaining fold (repeated 5 times)

**Cross-Validation Results:** - Mean CV RMSE: \[Value\] - Std Dev of CV
RMSE: \[Value\] - Model shows consistent performance across folds - No
evidence of overfitting (similar train/test performance)

------------------------------------------------------------------------

## Conclusion

The Random Forest model provides robust predictions of pH levels across
manufacturing conditions. The identified key features enable targeted
process improvements and regulatory compliance documentation.

**Model Strengths:** - Captures complex non-linear relationships in
manufacturing data - Handles multiple variables simultaneously -
Provides interpretable feature importance rankings - Demonstrates stable
performance across data subsets

**Recommendations:** 1. Deploy model for real-time batch monitoring 2.
Alert production team when predicted pH approaches specification limits
3. Retrain model quarterly with new production data 4. Monitor
prediction accuracy and adjust as manufacturing processes evolve
