---
title: "Student Data Exploratory Data Analysis"
author: "Data Analysis"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(readxl)
library(tidyverse)
library(skimr)
library(GGally)
library(kableExtra)
library(patchwork)
```

## Data Loading

```{r load-data}
# Load the StudentData
df <- read_excel("data/StudentData.xlsx")

# Display basic information
cat("Data loaded successfully!\n")
cat("Dimensions:", dim(df)[1], "rows,", dim(df)[2], "columns\n")
```

## Data Overview

```{r data-structure}
# Display structure
str(df)
```

```{r first-rows}
# Display first few rows
head(df) %>%
  kable(caption = "First 6 rows of Student Data") %>%
  kable_styling()
```

## Summary Statistics

```{r summary-stats}
# Using skim for comprehensive summary
skim(df)
```

```{r basic-summary}
# Numeric summary
summary(df) %>%
  kable(caption = "Basic Summary Statistics")
```

## Missing Data Analysis

```{r missing-data}
# Check for missing values
missing_summary <- data.frame(
  Column = names(df),
  Missing_Count = colSums(is.na(df)),
  Missing_Percentage = round(colSums(is.na(df)) / nrow(df) * 100, 2)
)

missing_summary %>%
  filter(Missing_Count > 0) %>%
  kable(caption = "Missing Values Summary") %>%
  kable_styling()

if(sum(missing_summary$Missing_Count) == 0) {
  cat("No missing values detected in the dataset!\n")
}
```

## Univariate Analysis

### Numeric Variables

```{r numeric-distributions}
# Get numeric columns
numeric_cols <- df %>% select(where(is.numeric)) %>% names()

# Create density plots for numeric variables
if(length(numeric_cols) > 0) {
  df %>%
    select(all_of(numeric_cols)) %>%
    pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
    ggplot(aes(x = value, fill = variable)) +
    geom_histogram(bins = 30, alpha = 0.7) +
    facet_wrap(~variable, scales = "free") +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(title = "Distribution of Numeric Variables",
         x = "Value", y = "Frequency")
}
```

```{r numeric-box-plots}
# Box plots for numeric variables
if(length(numeric_cols) > 0) {
  df %>%
    select(all_of(numeric_cols)) %>%
    pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
    ggplot(aes(x = variable, y = value, fill = variable)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~variable, scales = "free") +
    theme_minimal() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = "Box Plots of Numeric Variables",
         x = "Variable", y = "Value")
}
```

### Categorical Variables

```{r categorical-analysis}
# Get categorical columns
categorical_cols <- df %>% select(where(is.character) | where(is.factor)) %>% names()

if(length(categorical_cols) > 0) {
  for(col in categorical_cols) {
    cat("\n### ", col, "\n\n")

    # Count table
    counts <- df %>%
      group_by(!!sym(col)) %>%
      summarise(Count = n(), Percentage = round(n() / nrow(df) * 100, 2)) %>%
      arrange(desc(Count))

    print(kable(counts, caption = paste("Frequency of", col)))
  }
}
```

```{r categorical-bar-plots}
# Bar plots for categorical variables
if(length(categorical_cols) > 0) {
  plots <- list()

  for(i in seq_along(categorical_cols)) {
    col <- categorical_cols[i]

    plots[[i]] <- df %>%
      group_by(!!sym(col)) %>%
      summarise(count = n()) %>%
      ggplot(aes(x = reorder(!!sym(col), -count), y = count, fill = !!sym(col))) +
      geom_bar(stat = "identity", alpha = 0.7) +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "none") +
      labs(title = paste("Distribution of", col),
           x = col, y = "Count")
  }

  # Display plots
  if(length(plots) > 0) {
    wrap_plots(plots, ncol = 2)
  }
}
```

## Bivariate Analysis

### Numeric Variables Correlation

```{r correlation-matrix}
# Correlation matrix for numeric variables
if(length(numeric_cols) > 1) {
  cor_matrix <- df %>%
    select(all_of(numeric_cols)) %>%
    cor(use = "complete.obs")

  print(kable(cor_matrix, caption = "Correlation Matrix - Numeric Variables", digits = 3))
}
```

```{r correlation-plot}
# Visualization of correlations
if(length(numeric_cols) > 1) {
  df %>%
    select(all_of(numeric_cols)) %>%
    ggcorr(method = c("complete.obs", "pearson"),
           label = TRUE,
           label_size = 3,
           hjust = 0.75) +
    labs(title = "Correlation Heatmap - Numeric Variables")
}
```

### Relationships Between Variables

```{r scatter-plots}
# Create scatter plots for numeric variables
if(length(numeric_cols) > 1) {
  df %>%
    select(all_of(numeric_cols)) %>%
    ggpairs(lower = list(continuous = wrap("points", alpha = 0.3)),
            diag = list(continuous = wrap("barDiag", alpha = 0.5)),
            upper = list(continuous = wrap("cor", size = 3))) +
    theme_minimal() +
    labs(title = "Pairwise Relationships - Numeric Variables")
}
```

## Key Findings

```{r key-findings}
cat("### Data Summary\n\n")
cat("- **Total Records:** ", nrow(df), "\n")
cat("- **Total Features:** ", ncol(df), "\n")
cat("- **Numeric Variables:** ", length(numeric_cols), "\n")
cat("- **Categorical Variables:** ", length(categorical_cols), "\n")
cat("- **Missing Values:** ", sum(is.na(df)), "\n\n")

if(length(numeric_cols) > 0) {
  cat("### Numeric Variables Summary\n\n")
  numeric_summary <- df %>%
    select(all_of(numeric_cols)) %>%
    summarise(across(everything(),
                    list(min = min,
                         max = max,
                         mean = mean,
                         median = median,
                         sd = sd),
                    .names = "{.col}_{.fn}"))
  print(numeric_summary)
}
```

## Recommendations for Further Analysis

- Investigate relationships between key variables
- Perform statistical tests if needed
- Consider data transformations for skewed distributions
- Identify and handle outliers if appropriate
- Validate data quality assumptions
